Draft

### 2019-08-31 draft.tex
HoTT+CT+LLPO consistent

\usepackage{tikz-cd}
%\usepackage{answers}
\usepackage{biblatex}
\addbibresource{bibliography.bib}
\usepackage{setspace}
%\usepackage{graphicx}
\usepackage{enumitem}
%\usepackage{tikz-cd}
%\usepackage{todonotes}
%\usepackage{multicol}
%\usepackage{mathrsfs}
%\usepackage[margin=1in]{geometry} 
\usepackage{hyperref}% http://ctan.org/pkg/hyperref

\usepackage{amsmath}
\usepackage{cleveref}% http://ctan.org/pkg/cleveref
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{ textcomp }
%\usepackage{ wasysym }
\usepackage{bookmark}
%\usepackage{accents}
\usepackage[parfill]{parskip}

@@

\chapter{Modalities}
%Under the view of types as $(\infty, 1)$-categories, functions between them are functors. 
%We can then view the universe $\mathcal{U}$ itslef as a category. 
\begin{defn}
A \textbf{modality} is a function $\bigcirc : \mathcal{U} \rightarrow \mathcal{U}$ with the following properties.
\begin{enumerate}
	\item For every type $A$ we have a function $\eta_A^\bigcirc : A \rightarrow \bigcirc\; A$
	\item for every $A : \mathcal{U}$ and every type family $B : \bigcirc A \rightarrow \mathcal{U}$ we have a function
	$$\mathtt{ind}_\bigcirc : \Big( \prod_{a : A}\bigcirc (B (\eta_A^\bigcirc\; a)) \Big) \rightarrow \prod_{z : \bigcirc A} \bigcirc (B\; z)$$
	\item For every $f : \prod_{a : A} \bigcirc (B(\eta_A^\bigcirc\; a))$ there is a path $\mathtt{ind}_\bigcirc (f)(\eta_A^\bigcirc\; a) = f\; a$
	\item For all $z,z' : \bigcirc \; A$, the function $\eta_{z=z'}^\bigcirc : (z = z') \rightarrow \bigcirc (z = z')$ is an equivalence.

@@

@@

\begin{lem}\label{NisBnull}
$\mathbb{N}$ is $\lVert B \rVert$-null.
\end{lem}
\begin{proof}
Given $a : A$ and $f : \lVert B \rVert \; a \rightarrow \mathbb{N}$ we need to prove that there exists unique $f' : \mathbf{1} \rightarrow \mathbb{N}$ through which $f$ factors, in the sense that $f = f' \circ g$, where $g : C\; a \rightarrow \mathbf{1}$ is the sole inhabitant of its function space. 
Functions with $\mathbf{1}$ as their domain are constant. 
We therefore need to find some $z : \mathbb{N}$ so that $\lambda\; \_.\; z$ is $f'$. 
To that end, we will use the following instance of the Independence Principle
\begin{multline} \label{IPlhs}
\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	 \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbb{N}} \prod_{p : B\; a} f^*\; p = z \Big\rVert \bigg)
\end{multline}
\begin{multline} \label{IPrhs}
\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	  \Big\lVert \sum_{z : \mathbb{N}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \prod_{p : B\; a} f^*\; p = z \Big\rVert \bigg)
\end{multline}
where $f^*$ is the composition $f \circ C^B$, with $C^B$ being the canonical function $B\; a \rightarrow C\; a$. 
Note that the left hand side \ref{IPlhs} is provable. 
So we have a witness of the right hand side \ref{IPrhs}.
By providing the odd $s_1$ and even $s_2$ subsequences of $a$ to it, we get hold of $z_1 : \mathbb{N}$ and $z_2 : \mathbb{N} $ respectively. 
Equality on $\mathbb{N}$ is decidable, therefore $(z_1 = z_2) + (z_1 \neq z_2)$ is provable.

@@

\begin{lem}
If $C, D$ are $\lVert B \rVert$-null then so is $C + D$.
\end{lem}
\begin{proof}
We need to show that for $f : \lVert B\; a\rVert \rightarrow C+D$ there exists unique $f' : \mathbf{1} \rightarrow C+D$ such that $f = f' \circ e$, where $e : \lVert B\; a\rVert \rightarrow \mathbf{1}$ is the sole inhabitant of its function space. 
We define $$g : C + D \rightarrow \mathbf{2}$$ 
using components
$$g_C :\equiv \lambda\; \_.\; 0 : C \rightarrow \mathbf{2}$$
$$g_D :\equiv \lambda\; \_.\; 1 : D \rightarrow \mathbf{2}$$
Same as we did for \cref{NisBnull} we can lift a variant of IP, that will help us in our endeavor, from the model of Extensional Type Theory. 
This variant takes the following arguments in the form of a 4-product
\begin{align*}
&a: &A
\\
&\bar{f}: &B\; a \rightarrow C+D
\\ &\_ :
\begin{split}
\bigg( \prod_{\bar{s} : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} \bar{s}(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} \bar{s}(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	\quad \Big(\prod_{n : \mathbb{N}}\bar{s}\; n = 0 \Big) \rightarrow  \sum_{z : \mathbf{2}} \prod_{b : B\; a} g(\bar{f}\; b) = z  \bigg)
\end{split}
\\
&r : & \sum_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big)
\end{align*}
and returns
$$\Big\lVert \sum_{z : \mathbf{2}} \prod_{b : B\; a} g(\bar{f}\; b) = z \Big\rVert$$
We provide this with the arguments required, where $\bar{f}$ is $f$ composed with the truncation map $|*| : B\; a \rightarrow \lVert B\; a\rVert$, to get a witness 
$$\bar{z} : \sum_{z : \mathbf{2}} \prod_{b : B\; a} g(\bar{f}\; b) = z$$ 
Note that we dropped the truncation since our goal is a mere proposition, therefore we can act as if we have an actual $\bar{z}$. 
Using $\bar{z}$ we can prove that
$$y : \sum_{z : \mathbf{2}} \prod_{b : \lVert B\; a\rVert} g(f\; b) = z$$
We can prove that 
$$y= 0 + y = 1$$ 
We use induction on this coproduct to prove our goal. 
Without loss of generality, we argue only for the case of $y = 0$. 
As a first step towards providing a candidate for $f'$, we will try to construct some $h : \lVert B\; a\rVert \rightarrow C$ such that $\mathtt{inl} \circ h = f$. 
\begin{center}
\begin{tikzcd}
                                                             & \mathbf{1} \arrow[rdd, "f'", dashed, bend left] \arrow[d, "h'"', dashed]              &                    &            \\
                                                             & C \arrow[rd, "\mathtt{inl}"]                                                          &                    &            \\
B\; a \arrow[rr, "\bar{f}"', bend right=49] \arrow[r, "|*|"] & \lVert B\;a\rVert \arrow[r, "f"] \arrow[u, "h", dashed] \arrow[uu, "e", bend left=49] & C+D \arrow[r, "g"] & \mathbf{2}
\end{tikzcd}
\end{center}
Let $b : \lVert B\; a\rVert$. 
First, note that we can prove 
$$\zeta : \prod_{a: C+D} \Big(\big(\sum_{c : C} \mathtt{inl}\; c = a\big) + \big( \sum_{d:D} \mathtt{inr}\; d = a \big) \Big)$$
by induction on $C+D$. 
We then take cases on 
$$\zeta (f\; b) : \big(\sum_{c : C} \mathtt{inl}\; c = f\; b\big) + \big( \sum_{d:D} \mathtt{inr}\; d = f\; b \big)$$
\begin{itemize}
	\item If $\bar{c} : \sum_{c : C} \mathtt{inl}\; c = f\; b$, we define $h\; b :\equiv \bar{c}$.
	Clearly $\mathtt{inl} \circ h = f$.
	\item If $\sum_{d : D} \mathtt{inl}\; d = f\; b$, then $g(f\;b) = 1$ but since $y=0$ we also have $g(f\;b) = 0$. Contradiction. 
	This case is resolved by Ex Falso.
\end{itemize}
Since $C$ is $\lVert B \rVert$-null, there exists $h' : \mathbf{1} \rightarrow C$ such that $h = h' \circ e$. 
Observe that $\mathtt{inl} \circ h' : \mathbf{1} \rightarrow C+D$, furthermore $(\mathtt{inl} \circ h') \circ e= f$. 
We have found a valid candidate for $f'$, what is left is to show it's unique. 
% We reduce this to showing that $\sum_{f' : \mathbf{1} \rightarrow C+D} f' \circ e = f$ is a mere proposition.

>>>>

\usepackage{tikz-cd}
%\usepackage{answers}
\usepackage{biblatex}
\addbibresource{bibliography.bib}
\usepackage{setspace}
%\usepackage{graphicx}
\usepackage{enumitem}
%\usepackage{tikz-cd}
%\usepackage{todonotes}
%\usepackage{multicol}
%\usepackage{mathrsfs}
%\usepackage[margin=1in]{geometry} 
\usepackage{hyperref}% http://ctan.org/pkg/hyperref
\usepackage{ stmaryrd }
\usepackage{amsmath}
\usepackage{cleveref}% http://ctan.org/pkg/cleveref
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{ textcomp }
%\usepackage{ wasysym }
\usepackage{bookmark}
%\usepackage{accents}
\usepackage[parfill]{parskip}

@@

\chapter{Modalities}
%Under the view of types as $(\infty, 1)$-categories, functions between them are functors. 
%We can then view the universe $\mathcal{U}$ itslef as a category. 
\begin{defn}
A \textbf{modality} is a function $\bigcirc : \mathcal{U} \rightarrow \mathcal{U}$ with the following properties.
\begin{enumerate}
	\item For every type $A$ we have a function $\eta_A^\bigcirc : A \rightarrow \bigcirc\; A$
	\item for every $A : \mathcal{U}$ and every type family $B : \bigcirc A \rightarrow \mathcal{U}$ we have a function
	$$\mathtt{ind}_\bigcirc : \Big( \prod_{a : A}\bigcirc (B (\eta_A^\bigcirc\; a)) \Big) \rightarrow \prod_{z : \bigcirc A} \bigcirc (B\; z)$$
	\item For every $f : \prod_{a : A} \bigcirc (B(\eta_A^\bigcirc\; a))$ and every $a : A$, there is a path $\mathtt{ind}_\bigcirc (f)(\eta_A^\bigcirc\; a) = f\; a$
	\item For all $z,z' : \bigcirc \; A$, the function $\eta_{z=z'}^\bigcirc : (z = z') \rightarrow \bigcirc (z = z')$ is an equivalence.

@@

Nullification at a family of types is an example of a modality, as laid out in \cite{1706.07526}, where it is presented as a higher inductive type, complete with constructors and eliminators. 
To avoid listing all these trappings, we will instead look at the important properties $\mathcal{L}_B : \mathcal{U} \rightarrow \mathcal{U}$ should have to be considered a nullification operator.

It must have all the data required for it to be a modality.\\
For all $X : \mathcal{U}$, $\mathcal{L}_B(X)$ must be $B$\textbf{-null}.

@@

We can now assume that we have access to a model $\mathcal{E}$ of univalent type theory, in which CT, MP and IP' hold.

\begin{lem}\label{NisBnull}
$\mathbb{N}$ is $\lVert B \rVert$-null in $\mathcal{E}$.
\end{lem}
\begin{proof}
Given $a : A$ and $f : \lVert B \rVert \; a \rightarrow \mathbb{N}$ we need to prove that there exists unique $f' : \mathbf{1} \rightarrow \mathbb{N}$ through which $f$ factors, in the sense that $f = f' \circ g$, where $g : C\; a \rightarrow \mathbf{1}$ is the sole inhabitant of its function space. 
Functions with $\mathbf{1}$ as their domain are constant. 
We therefore need to find some $z : \mathbb{N}$ so that $\lambda\; \_.\; z$ is $f'$. 
To that end, we will use the following instance of the Independence Principle
\begin{multline} \label{IPlhs}
\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	 \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbb{N}} \prod_{p : B\; a} f^*\; p = z \Big\rVert \bigg)
\end{multline}
\begin{multline} \label{IPrhs}
\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	  \Big\lVert \sum_{z : \mathbb{N}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \prod_{p : B\; a} f^*\; p = z \Big\rVert \bigg)
\end{multline}
where $f^*$ is the composition $f \circ C^B$, with $C^B$ being the canonical function $B\; a \rightarrow C\; a$. 
Note that the left hand side \ref{IPlhs} is provable. 
So we have a witness of the right hand side \ref{IPrhs}.
By providing the odd $s_1$ and even $s_2$ subsequences of $a$ to it, we get hold of $z_1 : \mathbb{N}$ and $z_2 : \mathbb{N} $ respectively. 
Equality on $\mathbb{N}$ is decidable, therefore $(z_1 = z_2) + (z_1 \neq z_2)$ is provable.

@@

\section{CT + LLPO + HoTT}

We form a null types model $\mathcal{E}_{\lVert B \rVert}$ as defined in Section 5 of \cite{1905.03014} based on $\mathcal{E}$ from before. 
The types of this model are the $\lVert B \rVert $-null types of $\mathcal{E}$. 
By \cref{NisBnull} we have that $\llbracket \mathbb{N} \rrbracket^{\mathcal{E}_{\lVert B \rVert}} = \llbracket \mathbb{N} \rrbracket^\mathcal{E}$. 
Let $\mathcal{L}_{\lVert B \rVert} : \mathcal{U} \rightarrow \mathcal{U}$ be the $\lVert B \rVert$-nullification operator on $\mathcal{E}$. 
On $\mathcal{E}_{\lVert B \rVert}$, dependend products, dependent sums and identity on naturals are all inherited from $\mathcal{E}$, while propositional truncation $\llbracket \lVert \cdot \rVert \rrbracket^{\mathcal{E}_{\lVert B \rVert}}$ is $\mathcal{L}_{\lVert B \rVert} \lVert\cdot \rVert $ on $\mathcal{E}$.

We can then deduce that the interpretation of CT in $\mathcal{E}_{\lVert B \rVert}$ is the type 
$$\prod_{f : \mathbb{N}\rightarrow \mathbb{N}} \mathcal{L}_{\lVert B \rVert}\;\Big\lVert \sum_{e : \mathbb{N}} \prod_{x : \mathbb{N}} \sum_{z : \mathbb{N}} T(e,x,z) \times U(z) = f(x) \Big\rVert$$
in $\mathcal{E}$. 
To clarify, the types $T$ and $U$ remain the same because they are functions from naturals to naturals and we know that $\mathbb{N}$ is the same in both $\mathcal{E}_{\lVert B \rVert}$ and $\mathcal{E}$. 
It's trivial to construct an inhabitant, in $\mathcal{E}$, of the above type using a witness of CT, which we know to be true in $\mathcal{E}$.

Therefore, $\mathcal{E}_{\lVert B \rVert}$ is a model of univalent type theory where both CT and LLPO hold.

\section{B-nullification preserves coproducts}
\begin{lem}
If $C, D$ are $\lVert B \rVert$-null then so is $C + D$.
\end{lem}
\begin{proof}
We need to show that for $f : \lVert B\; a\rVert \rightarrow C+D$ there exists unique $f' : \mathbf{1} \rightarrow C+D$ such that $f = f' \circ e$, where $e : \lVert B\; a\rVert \rightarrow \mathbf{1}$ is the sole inhabitant of its function space. 
We define $$g : C + D \rightarrow \mathbf{2}$$ 
using components
$$g_C :\equiv \lambda\; \_.\; 0 : C \rightarrow \mathbf{2}$$
$$g_D :\equiv \lambda\; \_.\; 1 : D \rightarrow \mathbf{2}$$
Same as we did for \cref{NisBnull} we can lift a variant of IP, that will help us in our endeavor, from the model of Extensional Type Theory. 
This variant takes the following arguments in the form of a 4-product
\begin{align*}
&a: &A
\\
&\bar{f}: &B\; a \rightarrow C+D
\\ &\_ :
\begin{split}
\bigg( \prod_{\bar{s} : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} \bar{s}(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} \bar{s}(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	\quad \Big(\prod_{n : \mathbb{N}}\bar{s}\; n = 0 \Big) \rightarrow  \sum_{z : \mathbf{2}} \prod_{b : B\; a} g(\bar{f}\; b) = z  \bigg)
\end{split}
\\
&r : & \sum_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big)
\end{align*}
and returns
$$\Big\lVert \sum_{z : \mathbf{2}} \prod_{b : B\; a} g(\bar{f}\; b) = z \Big\rVert$$
We provide this with the arguments required, where $\bar{f}$ is $f$ composed with the truncation map $|*| : B\; a \rightarrow \lVert B\; a\rVert$, to get a witness 
$$\bar{z} : \sum_{z : \mathbf{2}} \prod_{b : B\; a} g(\bar{f}\; b) = z$$ 
Note that we dropped the truncation since our goal is a mere proposition, therefore we can act as if we have an actual $\bar{z}$. 
Using $\bar{z}$ we can prove that
$$y : \sum_{z : \mathbf{2}} \prod_{b : \lVert B\; a\rVert} g(f\; b) = z$$
We can prove that 
$$y= 0 + y = 1$$ 
We use induction on this coproduct to prove our goal. 
Without loss of generality, we argue only for the case of $y = 0$. 
As a first step towards providing a candidate for $f'$, we will try to construct some $h : \lVert B\; a\rVert \rightarrow C$ such that $\mathtt{inl} \circ h = f$. 
\begin{center}
\begin{tikzcd}
                                                             & \mathbf{1} \arrow[rdd, "f'", dashed, bend left] \arrow[d, "h'"', dashed]              &                    &            \\
                                                             & C \arrow[rd, "\mathtt{inl}"]                                                          &                    &            \\
B\; a \arrow[rr, "\bar{f}"', bend right=49] \arrow[r, "|*|"] & \lVert B\;a\rVert \arrow[r, "f"] \arrow[u, "h", dashed] \arrow[uu, "e", bend left=49] & C+D \arrow[r, "g"] & \mathbf{2}
\end{tikzcd}
\end{center}
Let $b : \lVert B\; a\rVert$. 
First, note that we can prove 
$$\zeta : \prod_{a: C+D} \Big(\big(\sum_{c : C} \mathtt{inl}\; c = a\big) + \big( \sum_{d:D} \mathtt{inr}\; d = a \big) \Big)$$
by induction on $C+D$. 
We then take cases on 
$$\zeta (f\; b) : \big(\sum_{c : C} \mathtt{inl}\; c = f\; b\big) + \big( \sum_{d:D} \mathtt{inr}\; d = f\; b \big)$$
\begin{itemize}
	\item If $\bar{c} : \sum_{c : C} \mathtt{inl}\; c = f\; b$, we define $h\; b :\equiv \bar{c}$.
	Clearly $\mathtt{inl} \circ h = f$.
	\item If $\sum_{d : D} \mathtt{inl}\; d = f\; b$, then $g(f\;b) = 1$ but since $y=0$ we also have $g(f\;b) = 0$. Contradiction. 
	This case is resolved by Ex Falso.
Since $C$ is $\lVert B \rVert$-null, there exists $h' : \mathbf{1} \rightarrow C$ such that $h = h' \circ e$. 
Observe that $\mathtt{inl} \circ h' : \mathbf{1} \rightarrow C+D$, furthermore $(\mathtt{inl} \circ h') \circ e= f$. 
We have found a valid candidate for $f'$, what is left is to show it's unique. 
% We reduce this to showing that $\sum_{f' : \mathbf{1} \rightarrow C+D} f' \circ e = f$ is a mere proposition.
